---
title: "MMCTime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MMCTime}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4, tibble.print_max = 4)
```

Load packages
```{r}
library(MMCTime)
library(ape)
library(ggtree)
library(dplyr)
set.seed(12345678)
```

This vignette demonstrates the basic usage of the MMCTime package.
We illustrate the use on a simulated phylogeny.

Simulate a genealogy from the Beta-coalescent
```{r,out.width="100%", dpi=300}
n_tips <- 100
samp_times <- 2020:2016
n_samp <- as.vector(rmultinom(1, n_tips, rep(1/5, 5)))

alpha <- 0.4
nu <- 1/4
sim <- simulate_beta(samp_times, n_samp, nu, alpha)

plot_mm_tree(sim$phy)+geom_tiplab(size=1.0)
print(sim$dates)
```

Simulate a mutation scaled phylogeny
```{r,out.width="100%", dpi=300}
phy_mut <- simulate_mut_arc(sim$phy, mu=1.0, omega=1.5)
ggtree(phy_mut) + geom_tiplab(size=1.0)
```

Time the phylogeny assuming the rooting is unknown, running 4 chains in parallel
We are using a rather low number of iterations here so that the vignette doesn't run for too long.
If the root position was known for example through the use of outgroup rooting, fix_root should be set to TRUE.
```{r}
res1 <- mmctime(phy_mut, sim$dates, model="beta", fix_root=FALSE, n_draws=5e2, thin=5e2) 
```
Print summaries. The Credible intervals cover simulation values rather well.
```{r}
print(res1$summaries, n=10)
```

Determine the tip order in the ground truth tree
```{r}
tip.order <- fortify(as.phylo(sim$phy), layout="rectangular") %>%
    dplyr::filter(.data$isTip) %>%
    dplyr::arrange(.data$y) %>%
    dplyr::pull("label")
```
First, plot a qualitative summary, using the ground truth tip order to enable easy comparison
```{r,fig.width=10, fig.height=10}
plot_densiCI(res1,tip.order=tip.order) + geom_tiplab()
```
Next, plot 16 posterior draws
```{r,fig.width=10, fig.height=10}
plot_mm_tree(sample_timetree(res1,16))
```
Finally, plot parameter pairs
```{r,fig.width=10, fig.height=10}
plot_pars(res1)
```