---
title: "MMCTime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MMCTime}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4, tibble.print_max = 4)
```

Load packages
```{r}
library(MMCTime)
library(ape)
library(ggtree)
library(dplyr)
library(ggplot2)
set.seed(12345678)
```

This vignette demonstrates the basic usage of the MMCTime package.
We illustrate the use on a simulated phylogeny.

## Simulations

Simulate a genealogy from the Beta-coalescent
```{r,out.width="100%", dpi=300}
n_tips <- 100
samp_times <- 2020:2016
n_samp_beta <- as.vector(rmultinom(1, n_tips, rep(1/5, 5)))

alpha <- 0.4
nu <- 1/4
sim_beta <- simulate_beta(samp_times, n_samp_beta, nu, alpha)

plot_mm_tree(sim_beta$phy)+geom_tiplab(size=1.0)
print(sim_beta$dates)
```

Simulate a genealogy from Kingman's coalescent
```{r,out.width="100%", dpi=300}
n_samp_km <- as.vector(rmultinom(1, n_tips, rep(1/5, 5)))

sim_km <- simulate_kingman(samp_times, n_samp_km, nu)

plot_mm_tree(sim_km$phy)+geom_tiplab(size=1.0)
print(sim_km$dates)
```

Simulate a mutation scaled phylogenies and plot them side by side
We can see that both have polytomies but visually the kingman tree has less uniform branch lengths.
```{r,out.width="45%", fig.show='hold', dpi=300}
phy_mut_beta <- simulate_mut_arc(sim_beta$phy, mu=1.0, omega=0.5)
phy_mut_km <- simulate_mut_arc(sim_km$phy, mu=1.0, omega=0.5)

ggtree(phy_mut_beta) + geom_tiplab(size=1.0) + ggtitle("Beta")
ggtree(phy_mut_km) + geom_tiplab(size=1.0) + ggtitle("Kingman")
```

## Inference

Time the phylogeny assuming the rooting is unknown, running 4 chains in parallel.

If the root position was known for example through the use of outgroup rooting, fix_root should be set to TRUE.

Note: This sort of analysis requires good temporal signal!

In practice you might want to run the chains for longer in case of low signal trees like these, especially the kingman tree.

```{r,eval=T}
res_beta <- mmctime(phy_mut_beta, sim_beta$dates, model="beta", fix_root=FALSE, n_draws=1e3, thin=1e3) 
res_km <- mmctime(phy_mut_km, sim_km$dates, model="beta", fix_root=FALSE, n_draws=1e3, thin=1e3) 
```

## Summaries and analysis

Print summaries for beta. The Credible intervals cover simulation values rather well.
```{r}
print(res_beta$summaries, n=10)
```
Same as above but for kingman
```{r}
print(res_km$summaries, n=10)
```
Plot the alpha parameter marginals.

We can clearly see that the alpha marginal concentrates around 1.0 for the kingman tree, i.e. the kingman limit of the beta coalescent.
```{r,out.width="100%", dpi=300}
alpha_df <- rbind(cbind(res_beta$draws[,c("alpha")],tree="beta"), cbind(res_km$draws[,c("alpha")],tree="kingman"))
ggplot(alpha_df) + geom_histogram(aes(x=alpha))+ facet_grid(rows=vars(tree))+theme_minimal()
```

We can further look at the posterior probability that the posterior tree contains no multiple mergers.

Since the q-variables indicate multiple mergers, we can simply sum these for each iterations and count the iterations where this sum equals 0. This gives the number of purely binary trees.

Observe that for the kingman posterior more than 1 in 12 trees is binary, whereas for the beta posterior it is less than 1 in a thousand.
```{r}
p_no_mm <- function(x)
{
  sum(apply(suppressWarnings(x$draws[,paste0("q_",1:199)]), 1, sum)==0)/x$n_draws
}

paste0("Posterior probability that a tree from the kingman posterior is binary: ", p_no_mm(res_km))
paste0("Posterior probability that a tree from the beta posterior is binary: ", p_no_mm(res_beta))
```

## Further Visualisation
First we aim to plot a qualitative densitree.

Determine the tip order in the ground truth tree.
```{r}
get_tip_ord <- function(x)
{
  fortify(as.phylo(x), layout="rectangular") %>%
      dplyr::filter(.data$isTip) %>%
      dplyr::arrange(.data$y) %>%
      dplyr::pull("label")
}
```
Plot a densitree for the beta posterior using the ground truth tip order to enable easy comparison.
```{r,fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
plot_densiCI(res_beta,tip.order=get_tip_ord(sim_beta$phy)) + geom_tiplab()
```
Same as above but for kingman tree.
```{r,fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
plot_densiCI(res_km,tip.order=get_tip_ord(sim_km$phy)) + geom_tiplab()
```
In practice what tends to happen is that for binary trees the multiple mergers fail to concentrate. 

We stress that this sort of setup is, however, designed for inference not model selection!

Next, plot 16 posterior draws for the beta posterior.
```{r,fig.width=10, fig.height=10}
plot_mm_tree(sample_timetree(res_beta,16))
```
And for the kingman posterior.
```{r,fig.width=10, fig.height=10}
plot_mm_tree(sample_timetree(res_km,16))
```
Plot parameter pairs for the beta posterior.
```{r,fig.width=10, fig.height=10}
plot_pars(res_beta)
```
Follwoed by the traces.
```{r,fig.width=10, fig.height=10}
plot_traces(res_beta)
```